Parsing Postfix Log Files

### 2009-08-24 doc/logparser-background.tex
Changes to capitalisation of titles, as specified by Howard Hamilton. Fixed that/than type identified by Joe Carthy.
Improve the description of Postfix restrictions, as specified by Howard Hamilton.

The sample database queries yield summary statistics about the efficiency of
anti-spam techniques.  Analysis of this kind is much easier to perform when
the data source is a database that when trying to directly analyse log
files.

@@

\section{Postfix}

@@

\subsection{Mixing And Matching Postfix Restrictions}

@@

Postfix uses one restriction list (containing zero or more restrictions)
for each stage of the \acronym{SMTP} conversation: client connection, HELO
command, MAIL FROM command, RCPT TO commands, DATA command, and end of
data.  The appropriate restriction list is evaluated for each stage, though
by default the restriction lists for client connection, HELO, and MAIL FROM
commands will not be evaluated until the first RCPT TO command is received,
because some clients do not deal properly with rejections before this
stage; a benefit of this delay is that Postfix has more information
available when logging rejections.

@@

Each restriction is evaluated to produce a result of \textit{reject},
\textit{permit}, \textit{dunno}, or the name of another restriction to be
evaluated; other less commonly used results are possible as described
in~\cite{access-man-page}.  The meaning of \textit{permit\/} and
\textit{reject\/} is obvious; \textit{dunno\/} means to stop evaluating the
current restriction and continue processing the remainder of the
restriction list, allowing exceptions to more general rules.  The
configuration example below shows how to require that all machines on the
local network except for the web server authenticate before sending mail;
the web server is exempt because the legacy applications running on it lack
authentication support.  The restriction list is evaluated from top to
bottom: \texttt{permit\_sasl\_authenticated} permits authenticated clients;
an exception is made for the web server in the
\texttt{check\_client\_access} check, and then other local machines are
rejected, because to continue past \texttt{permit\_sasl\_authenticated} the
client must not have authenticated.

@@

The administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex user-defined sequences
of lookups, restrictions, and exceptions.  Postfix tries to protect the
administrator from misconfiguration in as far as it reasonably can, e.g.\
the restriction \texttt{check\_helo\_mx\_access} cannot cause a mail to be
accepted, because the parameter it checks (the hostname given in the HELO
command) is under the control of the remote client.  Despite this, it is
possible for the administrator to make catastrophic mistakes, e.g.\
rejecting all mail --- the administrator must be cognisant of the effects
their configuration changes will have.  This is similar to one of UNIX's
design philosophies: \textit{``UNIX was not designed to stop its users from
doing stupid things, as that would also stop them from doing clever
things''\/}~\cite{unix-philosophy}.

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, or a rejection name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix, such as
allowing addresses associated with the payroll system to send mail on the
third Tuesday after pay day only, to help prevent problems from phishing
mails using faked sender addresses.  For example, a phishing mail might
pretend that the payroll system had a disastrous disk failure, and until
the server is replaced all salary payments will have to be processed
manually, so please reply to this mail with your name, address, and bank
account details; the criminal can then use any details sent to him to help
with identity theft.

>>>>

The sample database queries yield summary statistics about the efficiency of
anti-spam techniques.  Analysis of this kind is much easier to perform when
the data source is a database than when trying to directly analyse log
files.

@@

\section{The Postfix Mail Transfer Agent}

@@

\subsection{Mixing and Matching Postfix Restrictions}

@@

Postfix uses one restriction list (containing zero or more restrictions)
for each stage of the \acronym{SMTP} conversation: client connection, HELO
command, MAIL FROM command, RCPT TO commands, DATA command, and end of
data.  The appropriate restriction list is evaluated for each stage, though
by default the restriction lists for client connection, HELO, and MAIL FROM
commands will not be evaluated until the first RCPT TO command is received,
because some clients do not deal properly with rejections before this
stage; a benefit of this delay is that Postfix has more information
available when logging rejections.  Each restriction is evaluated to
produce a result of \textit{reject}, \textit{permit}, \textit{dunno}, or
the name of another restriction to be evaluated; other less commonly used
results are possible as described in~\cite{access-man-page}.  The meaning
of \textit{permit\/} and \textit{reject\/} is obvious; \textit{dunno\/}
means to stop evaluating the current restriction and continue processing
the remainder of the restriction list, allowing exceptions to more general
rules.  The administrator can define new restrictions as a list of existing
restrictions, allowing arbitrarily long and complex user-defined sequences
of lookups, restrictions, and exceptions.

@@

The configuration example below shows how to require that all machines on
the local network except for the web server authenticate before sending
mail; the web server is exempt because the legacy applications running on
it lack authentication support.  The restriction list is evaluated from top
to bottom: \texttt{permit\_sasl\_authenticated} permits authenticated
clients; an exception is made for the web server in the
\texttt{check\_client\_access} check, and then other local machines are
rejected, because to continue past \texttt{permit\_sasl\_authenticated} the
client must not have authenticated.

@@

Postfix tries to protect the administrator from misconfiguration in as far
as it reasonably can, e.g.\ the restriction
\texttt{check\_helo\_mx\_access} cannot cause a mail to be accepted,
because the parameter it checks (the hostname given in the HELO command) is
under the control of the remote client.  Despite this, it is possible for
the administrator to make catastrophic mistakes, e.g.\ rejecting all mail
--- the administrator must be cognisant of the effects their configuration
changes will have.  This is similar to one of UNIX's design philosophies:
\textit{``UNIX was not designed to stop its users from doing stupid things,
as that would also stop them from doing clever
things''\/}~\cite{unix-philosophy}.

@@

A policy server~\cite{policy-servers} is an external program consulted by
Postfix to determine the fate of an \acronym{SMTP} command.  The policy
server is given state information by Postfix (sample state information is
shown in \tableref{Example attributes sent to policy servers}) and returns
a result (reject, permit, dunno, or a restriction name) as described in
\sectionref{Mixing and matching Postfix restrictions}.  A policy server can
perform more complex checks than those provided by Postfix, such as
allowing addresses associated with the payroll system to send mail on the
third Tuesday after pay day only, to help prevent problems from phishing
mails using faked sender addresses.  For example, a phishing mail might
pretend that the payroll system had a disastrous disk failure, and until
the server is replaced all salary payments will have to be processed
manually, so please reply to this mail with your name, address, and bank
account details; the criminal can then use any details sent to him to help
with identity theft.
